---
title: "HELPmiss (Data Verbs) Activity"
author: "Abdulaziz Bokijonov"
output: html_notebook
---


## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData) # includes the `HELPmiss` data set 


# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")

```

## Helpful links: 

- Look though the DC Textbook for `tidyverse` functions. 
- Check out some of these RStudio cheat sheets:
  - <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf> 
  - <https://www.rstudio.com/resources/cheatsheets/>
  - <https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf>
  



## Task 1: Data Description

* The HELP study was a clinical trial for adult inpatients recruited from a detoxification unit.
* Patients with no primary care physician were randomized to receive a multidisciplinary assessment and a brief motivational intervention or usual care, with the goal of linking them to primary medical care.
* HELPmiss data includes data about drug addicted people and the data about their age, sex, substance, recovery and etc.
* This dataset is relatively filled, because it has very few not applicable data in it.
* Cases of this data are individuals that have drug problems.



## Task 2: Basic Summaries 

1. number of people (cases) in `HELPmiss` study

```{r}
#Your code here 
HELPmiss %>%
    summarise(total = n())
```

2. total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss`. 

```{r}
#Your code here 
HELPmiss %>%
    summarise(totalDetox = sum(e2b, na.rm=TRUE))
```


3. mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`


```{r}
#Your code here 
HELPmiss %>%
    summarise(mean = mean((subset(HELPmiss, anysub == "yes"))$daysanysub, na.rm=TRUE))
```



## Task 3: Group Summaries  

### males versus females
```{r}
#Your code here
HELPmiss %>%
    group_by(sex) %>%
    summarise(count = n()) %>%
    head()

```
From this command we can see that in total there were 359 men and 111 women in the study.

### homeless or not
```{r}
#Your code here
HELPmiss %>%
    group_by(homeless) %>%
    summarise(count = sum(e2b, na.rm=TRUE))%>%
    head()

```
Total number of times in the past 6 months entered a detox program (measured at baseline) for homeless people was higher than for housed.

### substance
```{r}
#Your code here
HELPmiss %>%
    group_by(substance) %>%
    summarise(count = n()) %>%
    head()

```
From this code we can see that 185 were on alcohol, 156 on cocaine, and 128 on heroin.

### homeless versus housed broken down by sex
```{r}
#Your code here
HELPmiss %>%
    group_by(sex, homeless) %>%
    summarise(count = n()) %>%
    head()

```
From this code we can see that there are 182 housed men and 69 women, and there are 177 homeless men and 42 women.

### homeless versus housed broken down by substance
```{r}
#Your code here
HELPmiss %>%
    group_by(substance, homeless) %>%
    summarise(count = n()) %>%
    head()
```
From this code we can see that alcoholics were more homeless than people on other substances.


## Task 4: Data Visualization & Observations

*Include one or more interesting plots from this data set involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can expand on the relationships that you studied in Task 2, or you can explore a different group of variables in `HELPmiss` that show something interesting. Remember to use the interactive commands in the console, generate the R commands that will reproduce your plot, and then paste the R commands into an R chunk in the RMarkdown file.*  

*Remember, you can use* `esquisser` *or* `mplot` *in your console. But only include the ggplot code in this Rmd document.*


```{r}
# Your Code here.
ggplot(HELPmiss) +
  aes(x = age, y = avg_drinks, colour = sex) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  theme_minimal()
```
From the plot with age on x-axis and average number of drinks consumed per day, in past 30 days, we can see that women, regardless of the age, usually consume more alcohol than men. Moreover, we can see that the amount of average drinks regardless of sex somewhat increases as the age increases.


